D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-13 ootavacuous.tex
February 13 eview of reworked material
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition above, and the term \emph{strict C++}\footnote{
	@@@ Future bikeshedding of name.}
% @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++.
to denote a hypothetical standard that excludes this prohibition, but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

Some workers recommend avoiding OOTA by ordering prior relaxed
loads before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this requires real instructions be executed, consuming real
time and real electrical power to solve a strictly theoretical
problem.
This might have been acceptable in the 1960s of my youth, but it is now
the year 2024.

@@

Goldblatt looked at interactions between OOTA cycles and
undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}, and
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes his examples and notes ways to separate UB and OOTA-cycle
concerns.
The remainder of this document will concentrate on examples lacking UB.

@@

Semantic dependencies are only one type of link in an OOTA cycle.
The other type extends from a given store to a load that returns the
value stored.
It is tempting to make the intuitive argument that the store must precede
the load in global time, and then to combine this with the notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% @@@ This citation credits the formulation to Ali Sezgin, third author
% @@@ of N4323 ("Out-of-Thin-Air Execution is Vacuous").
where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load ``reads-from'' links,
whether intra-thread
(rfi) or between threads (rfe).\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for definitions
	and properies of rf, rfe, and rfi.}

@@

@@@ Please see
Appendix~\ref{app:Informal Definition of Semantic Dependency}
for more a more detailed (albeit quite informal) definition.
Maybe eventually have an appendix that lists all the definitions
of both ``OOTA cycle'' and ``semantic dependency''.

@@

Consider the following:
\begin{quote}
\begin{verbatim}
x = y - z;
\end{verbatim}
\end{quote}
There appear to be semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation somehow knows that \co{y} is
always equal to \co{z} at this point then there is no semantic dependency;
the implementation can act as if the statement were simply ``\co{x = 0;}''.
We leave aside the question of how the implementation would know this,
given that \co{y} and \co{z} cannot be updated simultaneously
and are subject to change at any time by
other threads (a point we will return to in
section~\ref{sec:Global Optimization Can Destroy Dependencies}).

@@

Compare this example to the previous one:
\begin{quote}
\begin{verbatim}
if (x > 0)
    y = 42;
else
    y = 42;
\end{verbatim}
\end{quote}
Because the stores executed on each arm of the \co{if} statement write
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
for all intents and purposes a single store, independent of x.
Reasonable C++ implementations might disagree on this matter and
therefore on whether or not the example has a semantic dependency.
It is the implementations's choice.

@@

The implementation-defined aspects can affect whether or not an
abstract execution contains a semantic dependency.
As an example consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
for which \co{char} is a signed type,
but not those for which it is unsigned.

@@

(According to the current version of the standard, conflicting side effects
in unsequenced subexpressions constitute undefined behavior,
although there are proposals to make them defined in both C++ and C.\footnote{
	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}
Nevertheless, the example above is allowed because the order of evaluation
of arguments to a function call is ``indeterminately sequenced''
(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)

@@

For accesses to partially volatile objects, we remove some of the
requirements and add a new one.
We leave out the last two items; compilers are allowed to merge or
omit accesses to partially volatile objects.
And the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same object.
Accesses to different objects may be reordered relative to each other.
The new requirement is that the map must preserve the rf relation.
That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
omitted then the instruction it maps to in $H$ must read from the
instruction that $W$ maps to,
and consequently $W$ must not be omitted either.

@@

Consider this example, which appears to be an \co{if} statement:

@@

Consider the following:

@@

\emph{Any analysis of a given execution must either be fully informed
of the enclosing context, or much make conservative assumptions about
that context.}

@@

Consider this example, with \co{i} initially zero:

@@

Many projects' coding guidelines prohibit side effects in expressions,
their goal being to obtain portable code whose behavior does not depend
on arbitrary choices on the part of the implementation.
However, the above code nevertheless conforms to the standard.
Note that side effects in expressions lacking sequence points is
undefined behavior, however, there are proposals to define it in
the C++ and C languages, allowing implementations to nevertheless reorder
evaluation when permitted by the as-if rule.\footnote{
	\url{https://www.open-std.org/jtc1/sc22/wg21/docs/papers/2016/p0145r3.pdf}
	\url{https://www.open-std.org/jtc1/sc22/wg14/www/docs/n3203.htm}}

Compiler-based C++ implementions make their non-deterministic choices
at compile time, so that the object code is deterministic, at least
on a per-execution basis.

@@

The key point here is that diferent values might result in almost
completely disjoint executions.

Fortunately, there are now tools that can help.
For example, the SAT-solver-based \co{cbmc} tool
has been used to mechanically verify signicant portions of Linux-kernel
RCU from the C-language source
code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
Another tool, Nidhugg, which is based on partial-order
reduction, has been used to carry out a similar
verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
Both tools can easily check whether or not a given dependency is semantic
for at least some approximate definitions of ``semantic dependency''.
An example use of \co{cbmc} is illustrated in
Appendix~\ref{sec:Evaluating sdep Using cbmc}.

@@

The exact definition of a computer language is subject to some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each of which is subject to change over time.
Users  and implementers are of course wise to assume that their opinions
might be overridden by those of the standard and, for the users,
especially the implementation.
However, consider the following code:

@@

@@@ LTO?

@@

@@@  Need to add material on inventing loads, which is currently
in an appendix.

@@

Real-world constraints are imposed by the standard, which have been
considered in prior work.
However, in correct real-world C++ implementations, the standard's
abstract machine must be mapped onto real matter and energy.
This mapping has not been considered by prior work, and brings to
bear additional constraints from the laws of physics and from hardware
architecture and design, both of which have been only partially accounted
for in the standard.
One reason for the incomplete accounting for hardware constraints is
that accurate and executable formal descriptions of hardware memory
models did not appear until after the standard was released.
A final set of constraints is imposed by the C++ implementations
themselves, some of which have resulted in non-determinisms within
the standard.\footnote{
	For example, when pre-existing implementations could not agree
	on any specific deterministic handling of side-effects within
	expressions.
	On the other hand, implementer freedom does have significant
	value.}

@@

Plato's ``Timaeus'' notwithstanding, there has been much recent debate
as to whether the universe is in fact causal with cause necessarily
preceding effect.
However, no one has managed to construct a real time-travel machine
or any other apparatus that might result in effect preceding cause,
as discussed at length in Appendix~\ref{sec:Debates on Causality} and 
the URL that it provides.

@@

For more discussion of potential optimizations, please see
Section~\ref{sec:But What About Non-Volatile Atomics?}.

>>>>

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
prohibition above, and the term \emph{strict C++}\footnote{
	There is some discomfort with this name, so this footnote
	is reserved for future bikeshedding.}
% @@@ Alternatives thus far: OOTA-prone C++, OOTA-enabled C++,
% @@@ relaxed C++, OOTA C++, OOTA-agnostic C++, loose C++.
to denote a hypothetical standard that excludes this prohibition, but
is otherwise identical to full C++.
Unqualified \emph{C++} means full C++.

@@

Some workers recommend avoiding OOTA by ordering prior relaxed
loads before subsequent relaxed
stores~\cite{Boehm:2014:OGA:2618128.2618134,HansBoehm2019OOTArevisitedAgain,Lahav:2017:RSC:3062341.3062352},
but this can require real instructions be
executed~\cite[Section 7.1]{Maranget2012TutorialARMPower},
consuming real time and real electrical power to solve a strictly
theoretical problem.
This might have been acceptable in the 1960s of some of the authors'
youths, but it is now the year 2024.

@@

Goldblatt looked at interactions between OOTA cycles and
undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
% Appendix~\ref{app:Aside on Undefined Behavior}
% analyzes his examples and notes ways to separate UB and OOTA-cycle
% concerns.
The remainder of this document will concentrate on examples lacking UB.

@@

Semantic dependencies are only one type of link in an OOTA cycle.
The other type extends from a given store to a load that returns the
value stored.
It is tempting to make the intuitive argument that the store must precede
the load in global time, and then to combine this with the notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that OOTA cycles cannot occur.
The idea has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf~\cite{PaulEMcKenney2014OOTA},
% This citation credits the formulation to Ali Sezgin, third author
% of N4323 ("Out-of-Thin-Air Execution is Vacuous").
where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load ``reads-from'' links,
whether intra-thread
(rfi) or between threads (rfe).\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for definitions
	and properties of rf, rfe, and rfi.}

@@

% We no longer need this.  I will move A.1 to the cbmc appendix.
% Please see
% Appendix~\ref{app:Informal Definition of Semantic Dependency}
% for more a more detailed (albeit quite informal) definition.
% Maybe eventually have an appendix that lists all the definitions
% of both ``OOTA cycle'' and ``semantic dependency''.

@@

Consider the following:
\begin{quote}
\begin{verbatim}
x = y - z;
\end{verbatim}
\end{quote}
There appear to be semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation somehow knows that \co{y} is
always equal to \co{z} at this point then there is no semantic dependency;
the implementation can act as if the statement were simply ``\co{x = 0;}''.
We leave aside the question of how the implementation would know this,
given that \co{y} and \co{z} cannot be updated simultaneously
and are subject to change at any time by
other threads (a point we will return to in
section~\ref{sec:Global Optimization Can Destroy Dependencies}).
% The usual way in the Linux kernel is via a union, but it is probably
% best to ignore that possibility.

@@

Compare this example to the previous one:
\begin{quote}
\begin{verbatim}
if (x > 0)
    y = 42;
else
    y = 42;
\end{verbatim}
\end{quote}
Because the stores executed on each arm of the \co{if} statement write
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
for all intents and purposes a single store, independent of x.
Reasonable C++ implementations might disagree on this matter and
therefore on whether or not the example has a semantic dependency.
It is the implementation's choice.

@@

The implementation-defined aspects can affect whether or not an
abstract execution contains a semantic dependency.
As an example consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
running on implementations in
which \co{char} is a signed type,
but not those for which it is unsigned.

@@

(According to the current version of the standard, conflicting side effects
in unsequenced subexpressions constitute undefined behavior,
although there are proposals to make them defined in both
C++~\cite{GabrielDosReis2016P0145r3}
and C~\cite{AlexCeleste2023N3203}.
Nevertheless, the example above is allowed because the order of evaluation
of arguments to a function call is ``indeterminately sequenced''
(7.6.1.3p7 \co{[expr.call]}) rather than unsequenced, a subtle distinction.)

@@

For accesses to partially volatile objects, we remove some of the
requirements and add a new one.
We leave out the last two items; compilers are allowed to merge or
omit accesses to partially volatile objects.
And the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same object.
Accesses to different objects may be reordered relative to each other.
The new requirement is that the map must preserve the rf relation.
% Is this rf?  Or specifically rfe?
That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
omitted then the instruction it maps to in $H$ must read from the
instruction that $W$ maps to,
and consequently $W$ must not be omitted either.

Section~\ref{sec:Constraints of the Standard}
presents examples illustrating some of these requirements.

@@

Consider this example, similar to that shown in
Section~\ref{sec:Semantic Dependencies Not Affected by if Statements},
which appears to be an \co{if} statement:

@@

Consider the following example derived from that in
Section~\ref{sec:Semantic Dependencies and Source Code}:

@@

\emph{Any analysis of a given execution either must be fully informed
of the enclosing context, or must make conservative assumptions about
that context.}

@@

Consider this example, similar to that in
Section~\ref{sec:Abstract Executions},
with \co{i} initially zero:

@@

Many projects' coding guidelines prohibit side effects in expressions,
their goal being to obtain portable code whose behavior does not depend
on arbitrary choices on the part of the implementation.
However, the above code nevertheless conforms to the standard.
Note that side effects in expressions lacking sequence points is
undefined behavior, however, there are proposals to define it in the
C++~\cite{GabrielDosReis2016P0145r3} and
C~\cite{AlexCeleste2023N3203}
languages, allowing implementations to nevertheless reorder
evaluation when permitted by the as-if rule.

Even though the standard is non-deterministic, \emph{compiler-based C++
implementions make their non-deterministic choices at compile time, so
that the object code is deterministic, at least on a per-execution basis.}

@@

The key point here is that different values might result in almost
completely disjoint executions.

Fortunately, there are now tools that can help.
For example, the SAT-solver-based \co{cbmc} tool
has been used to mechanically verify signicant portions of Linux-kernel
RCU from the C-language source
code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
Another tool, Nidhugg, which is based on partial-order
reduction, has been used to carry out a similar
verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
Both tools can easily check whether or not a given dependency is semantic
for at least some approximate definitions of ``semantic dependency''.
% An example use of \co{cbmc} is illustrated in
% Appendix~\ref{sec:Evaluating sdep Using cbmc}.

@@

The exact definition of a computer language is subject to some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each of which is subject to change over time.
Users  and implementers are of course wise to assume that their opinions
might be overridden by those of the standard and, for the users,
especially the implementation.
However, consider the following code, similar to that shown in
Section~\ref{sec:Abstract Executions}:

@@

% @@@ LTO?  Does not appear to be relevant given new material.

@@

@@

Real-world constraints are imposed by the standard, which have been
considered in prior work.
However, in correct real-world C++ implementations, the standard's
abstract machine must be mapped onto real matter and energy.
This mapping has not been considered by prior work, and brings to
bear additional constraints from the laws of physics and from hardware
architecture and design, both of which have been only partially accounted
for in the standard and in prior work.
One reason for the incomplete accounting for hardware constraints is
that accurate and executable formal descriptions of hardware memory
models did not appear until after the standard was released.
A final set of constraints is imposed by the C++ implementations
themselves, some of which have resulted in non-determinisms within
the standard.\footnote{
	For example, when pre-existing implementations could not agree
	on any specific deterministic handling of side-effects within
	expressions.
	On the other hand, implementer freedom does have significant
	value.}

@@

Plato's ``Timaeus'' notwithstanding, there has been much recent debate
as to whether the universe is in fact causal with cause necessarily
preceding effect.
However, no one has managed to construct a real time-travel machine
or any other apparatus that might result in effect preceding cause.
% as discussed at length in Appendix~\ref{sec:Debates on Causality} and 
% the URL that it provides.

@@

For more discussion of potential optimizations, please see
Section~\ref{sec:But What About Non-Volatile Atomics?}.
% @@@ Pull that section in here?  Drop this forward reference?
